---
title: "Xgboost presentation"
output: 
  rmarkdown::html_vignette:
    css: vignette.css
    number_sections: yes
    toc: yes
bibliography: xgboost.bib
author: Tianqi Chen, Tong He, Michaël Benesty
vignette: >
  %\VignetteIndexEntry{Xgboost presentation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Introduction
============

This is an introductory document of using the \verb@xgboost@ package in *R*. 

**Xgboost** is short for e**X**treme **G**radient **B**oosting package. 

It is an efficient and scalable implementation of gradient boosting framework by @friedman2001greedy.

The package includes efficient *linear model* solver and *tree learning* algorithm. It supports various objective functions, including *regression*, *classification* and *ranking*. The package is made to be extendible, so that users are also allowed to define their own objectives easily. 

It has been [used](https://github.com/tqchen/xgboost) to win several [Kaggle](http://www.kaggle.com) competitions. 

It has several features:

* Speed: it can automatically do parallel computation on *Windows* and *Linux*, with *OpenMP*. It is generally over 10 times faster than `gbm`.
* Input Type: it takes several types of input data:
    * Dense Matrix: *R*'s dense matrix, i.e. `matrix` ;
    * Sparse Matrix: *R*'s sparse matrix, i.e. `Matrix::dgCMatrix` ;
    * Data File: local data files ;
    * `xgb.DMatrix`: it's own class (recommended) ;
* Sparsity: it accepts sparse input for both *tree booster*  and *linear booster*, and is optimized for sparse input ;
* Customization: it supports customized objective function and evaluation function ;
* Performance: it has better performance on several different datasets.

The purpose of this Vignette is to show you how to use **Xgboost** to make prediction from a model based on your own dataset.

Installation
============

For up-to-date version (which is *highly* recommended), please install from Github:

```{r installGithub, eval=FALSE}
devtools::install_github('tqchen/xgboost',subdir='R-package')
```

> *Windows* user will need to install [RTools](http://cran.r-project.org/bin/windows/Rtools/) first.

For stable version on CRAN, please run

```{r installCran, eval=FALSE}
install.packages('xgboost')
```

For the purpose of this tutorial we will load the required package.

```{r libLoading, results='hold', message=F, warning=F}
require(xgboost)
```

In this example, we are aiming to predict whether a mushroom can be eated (yeah, as always, example data are the exact one you will work on in your every day life :-). 

Mushroom data is cited from UCI Machine Learning Repository. @Bache+Lichman:2013.

Learning
========

Dataset loading
---------------

We will load the `agaricus` datasets embedded with the package and will link them to variables.

The datasets are already separated in `train` and `test` data:

* As their names imply, the `train` part will be used to build the model ;
* `test` will be used to check how well our model is. 

Without dividing the dataset we would test the model on data the algorithm have already seen. As you may imagine, it's not the best methodology to check the performance of a prediction (can it even be called a *prediction*?).

```{r datasetLoading, results='hold', message=F, warning=F}
data(agaricus.train, package='xgboost')
data(agaricus.test, package='xgboost')
train <- agaricus.train
test <- agaricus.test
# Each variable is a S3 object containing both label and data.
```

> In the real world, it would be up to you to make this division between `train` and `test` data. The way you should do it is out of the purpose of this article, however `caret` package may help.

The loaded `data` are stored in `dgCMatrix` which is a *sparse matrix* type and `label` is a `numeric` vector in `{0,1}`.

```{r dataClass, message=F, warning=F}
class(train$data)[1]
class(train$label)
```

`label` is the outcome of our dataset. It is the binary classification we want to predict in future data.

Basic Training using Xgboost
----------------------------

The most critical part of the process is the training one.

We are using the `train` data. As explained above, both `data` and `label` are in a variable.

In sparse matrix, cells which contains `0` are not encoded. Therefore, in a dataset where there are plenty of `0`, memory size is optimized. It is very usual to have such dataset. **Xgboost** can manage both dense and sparse matrix.

```{r trainingSparse, message=F, warning=F}
bstSparse <- xgboost(data = train$data, label = train$label, max.depth = 2, eta = 1, nround = 2, objective = "binary:logistic")
```

> To reach the value of a `S3` object field we use the `$` character.

Alternatively, you can put your dataset in a dense matrix, i.e. a basic R-matrix.

```{r trainingDense, message=F, warning=F}
bstDense <- xgboost(data = as.matrix(train$data), label = train$label, max.depth = 2, eta = 1, nround = 2,
               objective = "binary:logistic")
```

Above, data and label are not stored together. 

**Xgboost** offer a way to group them in a `xgb.DMatrix`. You can even add other meta data in it. It will be usefull for the most advanced features we will discover later.

```{r trainingDmatrix, message=F, warning=F}
dtrain <- xgb.DMatrix(data = train$data, label = train$label)
bstDMatrix <- xgboost(data = dtrain, max.depth = 2, eta = 1, nround = 2, objective = "binary:logistic")
```

**Xgboost** have plenty of features to help you to view how the learning progress internally. The obvious purpose is to help you to set the best parameters, which is the key of the quality of the model you are building.

One of the most simple way to see the progress is to set the `verbose` option. Look below of the effect of this parameter.

```{r trainingVerbose, message=T, warning=F}
# verbose 0, no message
bst <- xgboost(data = dtrain, max.depth = 2, eta = 1, nround = 2,
               objective = "binary:logistic", verbose = 0)

# verbose 1, print evaluation metric
bst <- xgboost(data = dtrain, max.depth = 2, eta = 1, nround = 2,
               objective = "binary:logistic", verbose = 1)

# verbose 2, also print information about tree
bst <- xgboost(data = dtrain, max.depth = 2, eta = 1, nround = 2,
               objective = "binary:logistic", verbose = 2)
```

Basic prediction using Xgboost
------------------------------

The main use of **Xgboost** is to predict data. For that purpose we will use the `test` dataset.

```{r predicting, message=F, warning=F}
pred <- predict(bst, test$data)
err <- mean(as.numeric(pred > 0.5) != test$label)
print(paste("test-error=", err))
```

> We remind you that the algorithm has never seen the `test` data.

Save and load models
--------------------

May be your dataset is big, and it takes time to train a model on it? May be you are not a big fan of loosing time in redoing the same task again and again? In these very rare cases, you will want to save your model and load it when required.

Hopefully for you, **Xgboost** implements such functions.

```{r saveModel, message=F, warning=F}
# save model to binary local file
xgb.save(bst, "xgboost.model")
```

An interesting test to see how identic our saved model is to the original one by comparing the two predictions.

```{r loadModel, message=F, warning=F}
# load binary model to R
bst2 <- xgb.load("xgboost.model")
pred2 <- predict(bst2, test$data)

# delete the created model (because cleaning is always better than dirtyness)
rm("xgboost.model")

# And now the test
print(paste("sum(abs(pred2-pred))=", sum(abs(pred2-pred))))
```

> result is `0`? We are good!

In some very specific cases, like when you want to pilot **Xgboost** from `caret`, you will want to save the model as a *R* binary vector. See below how to do it.

```{r saveLoadRBinVectorModel, message=F, warning=F}
# save model to R's raw vector
raw = xgb.save.raw(bst)

# load binary model to R
bst3 <- xgb.load(raw)
pred3 <- predict(bst3, test$data)

# pred2 should be identical to pred
print(paste("sum(abs(pred3-pred))=", sum(abs(pred2-pred))))
``` 

> Again `0`? It seems that `Xgboost` works prety well!

Advanced features
=================

Most of the features below have been created to help you to improve your model by offering a better understanding of its content.


Dataset preparation
-------------------

For the following advanced features, we need to put data in `xgb.DMatrix` as explained above.

```{r DMatrix, message=F, warning=F}
dtrain <- xgb.DMatrix(data = train$data, label=train$label)
dtest <- xgb.DMatrix(data = test$data, label=test$label)
```

Measure learning progress xgb.train
-----------------------------------

Both `xgboost` (simple) and `xgb.train` (advanced) functions train models.

One of the special feature of `xgb.train` is the capacity to follow the progress of the learning after each round. Because of the way boosting works, there is a time when having too many rounds lead to an overfitting. You can see this feature as a cousin of cross-validation method. The following features will help you to avoid overfitting or optimizing the learning time in stopping it as soon as possible.

One way to measure progress in learning of a model is to provide to the **Xgboost** a second dataset already classified. Therefore it can learn on the real dataset and test its model on the second one. Some metrics are measured after each round during the learning.

```{r watchlist, message=F, warning=F}
watchlist <- list(train=dtrain, test=dtest)

bst <- xgb.train(data=dtrain, max.depth=2, eta=1, nround=2, watchlist=watchlist,
                 objective = "binary:logistic")
```

> For the purpose of this example, we use `watchlist` parameter. It is a list of `xgb.DMatrix`, each of them tagged with a name.

For a better understanding of the learning progression, you may want to have some specific metric or even use multiple evaluation metrics.

```{r watchlist2, message=F, warning=F}
bst <- xgb.train(data=dtrain, max.depth=2, eta=1, nround=2, watchlist=watchlist,
                 eval.metric = "error", eval.metric = "logloss",
                 objective = "binary:logistic")
```

> `eval.metric` allows us to monitor the evaluation of several metrics at a time. Hereafter we will watch two new metrics, logloss and error.

Manipulating xgb.DMatrix
------------------------

### Save / Load

Like saving models, `xgb.DMatrix` object (which groups both dataset and outcome) can also be saved using `xgb.DMatrix.save` function.

```{r DMatrixSave, message=F, warning=F}
xgb.DMatrix.save(dtrain, "dtrain.buffer")
# to load it in, simply call xgb.DMatrix
dtrain2 <- xgb.DMatrix("dtrain.buffer")
bst <- xgb.train(data=dtrain2, max.depth=2, eta=1, nround=2, watchlist=watchlist,
                 objective = "binary:logistic")
```

### Information extraction

Information can be extracted from `xgb.DMatrix` using `getinfo` function. Hereafter we will extract `label` data.

```{r getinfo, message=F, warning=F}
label = getinfo(dtest, "label")
pred <- predict(bst, dtest)
err <- as.numeric(sum(as.integer(pred > 0.5) != label))/length(label)
print(paste("test-error=", err))
```

View the trees from a model
---------------------------

You can dump the tree you learned using `xgb.dump` into a text file.

```{r dump, message=T, warning=F}
xgb.dump(bst, with.stats = T)
```

> if you provide a path to `fname` parameter you can save the trees to your hard drive.

Feature importance
------------------

Finally, you can check which features are the most important and plot the result (more information in vignette [Discover your data with **Xgboost**](www.somewhere.com)).

```{r featureImportance, message=T, warning=F, fig.width=8, fig.height=5, fig.align='center'}
importance_matrix <- xgb.importance(feature_names = train$data@Dimnames[[2]], model = bst)
print(importance_matrix)
xgb.plot.importance(importance_matrix)
```

References
==========